Aligning marine species range data to better serve science and conservation

### 2016-03-10 data_explore/data_explore.Rmd
updates to manuscript based on CS and JA comments

### define 25%, 50%, and 75% quartiles for both x and y axes
area_align_med <- median(spp_list$area_ratio, na.rm = TRUE)
dist_align_med <- median(spp_list$sm_perc, na.rm = TRUE)

@@

### mongo plot time
scatter_arearatio_vs_overlap <- ggplot(spp_list %>% 
                                         transform(popn_category = reorder(popn_category, category_score)),
                                       aes(x = area_ratio, 
                                           y = sm_perc, 
                                           color = popn_category,
                                           size = lg_area_pct)) + 
  ggtheme_plot + 
  ### color the quadrant backgrounds:
    annotate("rect", xmin = area_align_med, xmax = 100, 
                     ymin = dist_align_med, ymax = 100, 
             alpha = .3, 
#             fill= "green4")  + 
             fill= "#4dac26")  + 
    annotate("rect", xmax = area_align_med, xmin =   0, 
                     ymin = dist_align_med, ymax = 100, 
             alpha = .3, 
#             fill= "greenyellow") + 
             fill= "#b8e186") + 
    annotate("rect", xmin = area_align_med, xmax = 100, 
                     ymax = dist_align_med, ymin =   0, 
             alpha = .3, 
#             fill= "orange") + 
             fill= "#f1b6da") + 
    annotate("rect", xmax = area_align_med, xmin =   0, 
                     ymax = dist_align_med, ymin =   0, 
             alpha = .3, 
#             fill= "orangered2") + 
             fill= "#d01c8b") + 
  geom_point(alpha = .6)

### Manage scales for color and size 
scatter_arearatio_vs_overlap <- scatter_arearatio_vs_overlap +
  scale_colour_manual(values = c('LC' = 'green4', 
                                 'NT' = 'greenyellow', 
                                 'VU' = 'yellow', 
                                 'EN' = 'orange', 
                                 'CR' = 'orangered2',
                                 'DD' = 'grey30')) + 
  scale_size_continuous(limits = c(0, 100),
                        breaks = c( 0,    25,    50,    75,    100), 
                        labels = c('0%', '25%', '50%', '75%', '100%')) + 
  scale_x_continuous(expand = c(0, 0), 
                     limits = c(0, 102),
                     breaks = c(seq(0, 100, 25))) +
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0, 102),
                     breaks = c(seq(0, 100, 25)))

### here are median labels:
scatter_arearatio_vs_overlap <- scatter_arearatio_vs_overlap +
  annotate(geom = 'text',
           x = area_align_med, y = 5,
           hjust = 0, vjust = 0,
           color = 'grey30', 
           size = 2,
           fontface = 'bold.italic', angle = 90,
           label = sprintf('Median = %s%%', round(area_align_med, 1))) +
  annotate(geom = 'text',
           x = 5, y = dist_align_med,
           hjust = 0, vjust = 0,
           color = 'grey30', 
           size = 2,
           fontface = 'bold.italic', angle = 0,
           label = sprintf('Median = %s%%', round(dist_align_med, 1)))

### here are median lines:
# scatter_arearatio_vs_overlap <- scatter_arearatio_vs_overlap +
#   geom_vline(x = area_align_med,
#              color = 'grey50', linetype = 'dashed') +
#   geom_hline(y = dist_align_med,
#              color = 'grey50', linetype = 'dashed')

### here are individual species, identified and named on the plot:
# scatter_arearatio_vs_overlap <- scatter_arearatio_vs_overlap +
#   geom_point(data = spp_list_label, shape = 4, color = 'grey30', size = 4) + 
#   geom_text(data = spp_list_label, 
#             aes(label = sciname_label), 
#             color = 'grey30',
#             fontface = 'bold.italic',
#             size = 2,
#             hjust = .25,
#             vjust = 1.5)

scatter_arearatio_vs_overlap <- scatter_arearatio_vs_overlap +
  labs(x = bquote('Extent alignment '*alpha['ext']*' (%)'), 
       y = bquote('Distribution alignment '*alpha['dist']*' (%)'), 
#       title = 'Distribution alignment vs Extent alignment',
       color = 'Risk', size = '% ocean range')

@@

### Create species lists by quadrant
spp_list_q1 <- spp_list1 %>%
  filter(area_ratio >= median(area_ratio) & sm_perc >= median(sm_perc))
spp_list_q2 <- spp_list1 %>%
  filter(area_ratio < median(area_ratio) & sm_perc >= median(sm_perc))
spp_list_q3 <- spp_list1 %>%
  filter(area_ratio >= median(area_ratio) & sm_perc < median(sm_perc))
spp_list_q4 <- spp_list1 %>%
  filter(area_ratio < median(area_ratio) & sm_perc < median(sm_perc))

@@

### Plot the bar chart
barchart_spp_gp_quads <- ggplot(spp_gp_quadrants %>% 
                                  transform(spp_group_text = reorder(spp_group_text, pct_q1)), 
                                aes(x = spp_group_text, fill = quad, weight = pct_quad)) +
  ggtheme_plot + 
  theme(panel.grid.major.y = element_blank()) +
  geom_bar(stat = 'count', alpha = .8) +
  scale_fill_manual(values = c('q1 well-matched'   = '#4dac26', 
                               'q2 dist-matched'   = '#b8e186', 
                               'q3 area-matched'   = '#f1b6da', 
                               'q4 poorly matched' = '#d01c8b')) +
#   scale_fill_manual(values = c('q1 well-matched'   = 'green4', 
#                                'q2 dist-matched'   = 'greenyellow', 
#                                'q3 area-matched'   = 'orange', 
#                                'q4 poorly matched' = 'orangered2')) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1.1)) +
  geom_text(aes(label = sprintf('n = %s', n_spp), y = 1.01), hjust = 0, 
            size = 2, 
            color = 'grey30') +
  coord_flip() +
  labs(x = 'Taxonomic Group', 
       y = 'Relative proportion of species per quadrant', 
#       title = 'Taxonomic group by quadrant',
       fill = 'Alignment')

@@

``` {r examine_risk_quartiles, echo = FALSE, eval = FALSE}

@@

### Plot the bar chart
barchart_spp_risk_quads <- ggplot(spp_risk_quadrants1 %>% 
                                  transform(popn_category = reorder(popn_category, cat_rank)), 
                                aes(x = popn_category, fill = quad, weight = pct_quad)) +
  ggtheme_plot + 
  theme(panel.grid.major.y = element_blank()) +
  geom_bar(stat = 'count', alpha = .8) +
  scale_fill_manual(values = c('q1 well-matched'   = '#4dac26', 
                               'q2 dist-matched'   = '#b8e186', 
                               'q3 area-matched'   = '#f1b6da', 
                               'q4 poorly matched' = '#d01c8b')) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1.1)) +
  geom_text(aes(label = sprintf('n = %s', n_tot_cat), y = 1.01), hjust = 0, 
            size = 2, 
            color = 'grey30') +
  coord_flip() +
  labs(x = 'Risk category', 
       y = 'Relative proportion of species per quadrant', 
#       title = 'Risk category by quadrant',
       fill = 'Alignment')

@@

``` {r plot bar charts by taxa, echo = FALSE, message = FALSE, warning = FALSE, eval = FALSE}
### bar charts comparing number of species in each class for each data set

@@

bar_taxa2 <- ggplot(taxa_df_long2, aes(x = class, y = percent, fill = dataset)) +
  ggtheme_plot + 
  theme(panel.grid.major.x = element_blank()) +
  geom_bar(stat = 'identity', alpha = .8) +
#  scale_fill_manual(values = c('AquaMaps' = 'red', 'Both' = 'mediumorchid', 'IUCN' = 'blue')) +
  scale_fill_manual(values = c('AquaMaps' = '#1b9e77', 'Both' = '#d95f02', 'IUCN' = '#7570b3')) +
  geom_text(aes(label = sprintf('n = %s', n_total), y = 102), 
            size = 2, 
            angle = 90, hjust = 0, color = 'grey30') +
  theme(axis.text.x = element_text(angle = 75, hjust = 1)) + 
  scale_y_continuous(expand = c(0, 0), 
                     limits = c(0, 130),
                     breaks = c(0, 25, 50, 75, 100)) + 
  labs(# title = 'Species maps by taxonomic class',
       x = 'Class', 
       y = 'Percent of mapped species',
       fill = 'Dataset')
print(bar_taxa2)
ggsave(filename = file.path(dir_fig, 'bar_by_taxa_all.png'), 
       width = 17.8, height = 7, units = 'cm')

@@

``` {r get_iucn_citation_info, echo = FALSE}

>>>>

### define 25%, 50%, and 75% quartiles for both x and y axes
area_align_mean <- mean(spp_list$area_ratio, na.rm = TRUE)
dist_align_mean <- mean(spp_list$sm_perc, na.rm = TRUE)

@@

### mongo plot time
scatter_arearatio_vs_overlap <- ggplot(spp_list %>% 
                                         transform(popn_category = reorder(popn_category, category_score)),
                                       aes(x = area_ratio, 
                                           y = sm_perc, 
                                           color = popn_category,
                                           size = lg_area_pct)) + 
  ggtheme_plot + 
  ### color the quadrant backgrounds:
    annotate("rect", xmin = area_align_mean, xmax = 100, 
                     ymin = dist_align_mean, ymax = 100, 
             alpha = .3, 
#             fill= "green4")  + 
             fill= "#4dac26")  + 
    annotate("rect", xmax = area_align_mean, xmin =   0, 
                     ymin = dist_align_mean, ymax = 100, 
             alpha = .3, 
#             fill= "greenyellow") + 
             fill= "#b8e186") + 
    annotate("rect", xmin = area_align_mean, xmax = 100, 
                     ymax = dist_align_mean, ymin =   0, 
             alpha = .3, 
#             fill= "orange") + 
             fill= "#f1b6da") + 
    annotate("rect", xmax = area_align_mean, xmin =   0, 
                     ymax = dist_align_mean, ymin =   0, 
             alpha = .3, 
#             fill= "orangered2") + 
             fill= "#d01c8b") + 
  geom_point(alpha = .6)

### Manage scales for color and size 
scatter_arearatio_vs_overlap <- scatter_arearatio_vs_overlap +
  scale_colour_manual(values = c('LC' = 'green4', 
                                 'NT' = 'greenyellow', 
                                 'VU' = 'yellow', 
                                 'EN' = 'orange', 
                                 'CR' = 'orangered2',
                                 'DD' = 'grey30')) + 
  scale_size_continuous(limits = c(0, 100),
                        breaks = c( 0,    25,    50,    75,    100), 
                        labels = c('0%', '25%', '50%', '75%', '100%')) + 
  scale_x_continuous(expand = c(0, 0), 
                     limits = c(0, 102),
                     breaks = c(seq(0, 100, 25)),
                     labels = c('0%', '25%', '50%', '75%', '100%')) +
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0, 102),
                     breaks = c(seq(0, 100, 25)),
                     labels = c('0%', '25%', '50%', '75%', '100%'))

### here are mean labels:
scatter_arearatio_vs_overlap <- scatter_arearatio_vs_overlap +
  annotate(geom = 'text',
           x = area_align_mean, y = 5,
           hjust = 0, vjust = 0,
           color = 'grey30', 
           size = 2,
           fontface = 'bold.italic', angle = 90,
           label = sprintf('Mean = %s%%', round(area_align_mean, 1))) +
  annotate(geom = 'text',
           x = 5, y = dist_align_mean,
           hjust = 0, vjust = 0,
           color = 'grey30', 
           size = 2,
           fontface = 'bold.italic', angle = 0,
           label = sprintf('Mean = %s%%', round(dist_align_mean, 1)))

### here are mean lines:
# scatter_arearatio_vs_overlap <- scatter_arearatio_vs_overlap +
#   geom_vline(x = area_align_mean,
#              color = 'grey50', linetype = 'dashed') +
#   geom_hline(y = dist_align_mean,
#              color = 'grey50', linetype = 'dashed')

### here are individual species, identified and namean on the plot:
# scatter_arearatio_vs_overlap <- scatter_arearatio_vs_overlap +
#   geom_point(data = spp_list_label, shape = 4, color = 'grey30', size = 4) + 
#   geom_text(data = spp_list_label, 
#             aes(label = sciname_label), 
#             color = 'grey30',
#             fontface = 'bold.italic',
#             size = 2,
#             hjust = .25,
#             vjust = 1.5)

scatter_arearatio_vs_overlap <- scatter_arearatio_vs_overlap +
  labs(x = bquote('Area alignment '*alpha['ext']*' (%)'), 
       y = bquote('Distribution alignment '*alpha['dist']*' (%)'), 
#       title = 'Distribution alignment vs Extent alignment',
       color = 'Risk', size = '% ocean range')

@@

### Create species lists by quadrant
spp_list_q1 <- spp_list1 %>%
  filter(area_ratio >= mean(area_ratio) & sm_perc >= mean(sm_perc))
spp_list_q2 <- spp_list1 %>%
  filter(area_ratio < mean(area_ratio) & sm_perc >= mean(sm_perc))
spp_list_q3 <- spp_list1 %>%
  filter(area_ratio >= mean(area_ratio) & sm_perc < mean(sm_perc))
spp_list_q4 <- spp_list1 %>%
  filter(area_ratio < mean(area_ratio) & sm_perc < mean(sm_perc))

@@

### Plot the bar chart
barchart_spp_gp_quads <- ggplot(spp_gp_quadrants %>% 
                                  transform(spp_group_text = reorder(spp_group_text, pct_q1)), 
                                aes(x = spp_group_text, fill = quad, weight = pct_quad)) +
  ggtheme_plot + 
  theme(panel.grid.major.y = element_blank()) +
  geom_bar(stat = 'count', alpha = .8) +
  scale_fill_manual(values = c('q1 well-matched'   = '#4dac26', 
                               'q2 dist-matched'   = '#b8e186', 
                               'q3 area-matched'   = '#f1b6da', 
                               'q4 poorly matched' = '#d01c8b')) +
#   scale_fill_manual(values = c('q1 well-matched'   = 'green4', 
#                                'q2 dist-matched'   = 'greenyellow', 
#                                'q3 area-matched'   = 'orange', 
#                                'q4 poorly matched' = 'orangered2')) +
  scale_y_continuous(expand = c(0, 0), 
                     limits = c(0, 1.1),
                     breaks = c(0, .25, .5, .75, 1.0),
                     labels = c('0%', '25%', '50%', '75%', '100%')) +
  geom_text(aes(label = sprintf('n = %s', n_spp), y = 1.01), hjust = 0, 
            size = 2, 
            color = 'grey30') +
  coord_flip() +
  labs(x = 'Taxonomic Group', 
       y = 'Percent of species by quadrant', 
#       title = 'Taxonomic group by quadrant',
       fill = 'Alignment')

@@

``` {r count_occurrences_by_quad, echo = FALSE, warning = FALSE}

gbif_occur <- read_csv(file.path(dir_data, 'gbif_occurrence.csv'))
obis_occur <- read_csv(file.path(dir_data, 'obis_occurrence.csv'))

totes_occur <- obis_occur %>%
  left_join(gbif_occur) %>%
  rowwise() %>%
  mutate(mean_occ = mean(c(gbif_occ, obis_occ), na.rm = TRUE)) %>%
  group_by(iucn_sid) %>%
  summarize(mean_occ = mean(mean_occ, na.rm = TRUE),
            tots_occ = sum(mean_occ, na.rm = TRUE))

spp_quads_occ <- spp_list_quads %>%
  left_join(totes_occur %>%
              dplyr::select(iucn_sid, mean_occ),
            by = 'iucn_sid')

write_csv(spp_quads_occ, file.path(dir_data, 'spp_quads_occ.csv'))

``` {r examine_risk_quartiles, echo = FALSE, eval = TRUE}

@@

### Plot the bar chart
barchart_spp_risk_quads <- ggplot(spp_risk_quadrants1 %>% 
                                  transform(popn_category = reorder(popn_category, cat_rank)), 
                                aes(x = popn_category, fill = quad, weight = pct_quad)) +
  ggtheme_plot + 
  theme(panel.grid.major.y = element_blank()) +
  geom_bar(stat = 'count', alpha = .8) +
  scale_fill_manual(values = c('q1 well-matched'   = '#4dac26', 
                               'q2 dist-matched'   = '#b8e186', 
                               'q3 area-matched'   = '#f1b6da', 
                               'q4 poorly matched' = '#d01c8b')) +
  scale_y_continuous(expand = c(0, 0), 
                     limits = c(0, 1.1),
                     breaks = c(0, .25, .5, .75, 1.0),
                     labels = c('0%', '25%', '50%', '75%', '100%')) +
  geom_text(aes(label = sprintf('n = %s', n_tot_cat), y = 1.01), hjust = 0, 
            size = 2, 
            color = 'grey30') +
  coord_flip() +
  labs(x = 'Risk category', 
       y = 'Percent of species by quadrant', 
#       title = 'Risk category by quadrant',
       fill = 'Alignment')

@@

``` {r plot bar charts by taxa, echo = FALSE, message = FALSE, warning = FALSE, eval = TRUE}
### bar charts comparing number of species in each class for each data set

@@

bar_taxa2 <- ggplot(taxa_df_long2, aes(x = class, y = percent, fill = dataset)) +
  ggtheme_plot + 
  theme(panel.grid.major.x = element_blank()) +
  geom_bar(stat = 'identity', alpha = .8) +
#  scale_fill_manual(values = c('AquaMaps' = 'red', 'Both' = 'mediumorchid', 'IUCN' = 'blue')) +
  scale_fill_manual(values = c('AquaMaps' = '#1b9e77', 'Both' = '#d95f02', 'IUCN' = '#7570b3')) +
  geom_text(aes(label = sprintf('n = %s', n_total), y = 102), 
            size = 2, 
            angle = 90, hjust = 0, color = 'grey30') +
  theme(axis.text.x = element_text(angle = 75, hjust = 1)) + 
  scale_y_continuous(expand = c(0, 0), 
                     limits = c(0, 130),
                     breaks = c(0, 25, 50, 75, 100),
                     labels = c('0%', '25%', '50%', '75%', '100%')) + 
  labs(# title = 'Species maps by taxonomic class',
       x = 'Class', 
       y = 'Percent of mapped species',
       fill = 'Dataset')
print(bar_taxa2)
ggsave(filename = file.path(dir_fig, 'bar_by_taxa_all.png'), 
       width = 17.8, height = 7, units = 'cm')

@@

``` {r get_iucn_citation_info, echo = FALSE, warning = FALSE, eval = FALSE}

### 2016-03-10 manuscript-main.Rmd
updates to manuscript based on CS and JA comments

>>>>

``` {r calc species numbers, echo = FALSE}
n_spp_am     <- nrow(spp_am)
n_spp_am_rev <- nrow(spp_am %>% filter(reviewed == 1))
n_spp_iucn   <- length(unique(spp_iucn_maps$iucn_sid))
n_spp_both   <- length(unique(spp_map_pairs$iucn_sid))
n_spp_all    <- n_spp_am + n_spp_iucn - n_spp_both

p_spp_am     <- round(n_spp_am/n_spp_all * 100, 1)
p_spp_iucn   <- round(n_spp_iucn/n_spp_all * 100, 1)
p_both_am    <- round(n_spp_both/n_spp_am   * 100, 1)
p_both_iucn  <- round(n_spp_both/n_spp_iucn * 100, 1)
p_both_all   <- round(n_spp_both/n_spp_all  * 100, 1)

spp_list <- read_csv(file.path(dir_data, 'spp_list_w_area_trimmed.csv'))
area_align_mean <- round(mean(spp_list$area_ratio, na.rm = TRUE), 1)
dist_align_mean <- round(mean(spp_list$sm_perc, na.rm = TRUE), 1)

```
-----

### 2016-03-10 ms_1_abst_sig.Rmd
updates to manuscript based on CS and JA comments

Species distribution data provide the foundation for a wide range of ecological research and conservation science and management, yet most species ranges remain unknown, and existing range maps have little overlap. In the ocean, two global species distribution datasets, produced by AquaMaps (REF) and the International Union for Conservation of Nature (IUCN) (REF), dominate our understanding of marine species ranges throughout the world's oceans.  Together they represent `r format(n_spp_all, big.mark = ',')` species, mostly from AquaMaps, with only `r format(n_spp_both, big.mark = ',')` species overlapping.  Here we examine differences in predicted species ranges between the two datasets, propose mechanistic causes and potential solutions for such differences, and explore the implications of these differences for management and conservation decisions.  We find that IUCN maps often disregard bathymetry for depth-limited species, leading to predictions of species presence at unsuitable depths, and AquaMaps ranges for data-poor species often extrapolate presence far afield from known occurrences.  We illustrate the implications of these differences by repeating two recent applications - the Ocean Health Index (Halpern et al. 2012, 2015) biodiversity goal and a global analysis of gaps in coverage of marine protected areas (Klein et al. 2015) - and find significantly different estimates of the status of global biodiversity and effectiveness of conservation depending on how the two datasets were prioritized. ___Creating a single, highly accurate dataset of global marine species ranges will be difficult. Understanding the implications of dataset differences for conservation planning and decision-making is essential.___

>>>>

Species distribution data provide the foundation for a wide range of ecological research studies and conservation management decisions, yet most species ranges remain unknown, and existing range maps often suffer from data limitations and show inconsistencies.  AquaMaps (Kaschner et al. ___2015???___) and the International Union for Conservation of Nature (IUCN) (IUCN 2015) are two distinct efforts to map marine species distributions at a global scale.  Together these two databases represent `r format(n_spp_all, big.mark = ',')` species (`r format(p_spp_am, big.mark = ',')`% within AquaMaps, `r format(p_spp_iucn, big.mark = ',')`% within IUCN), with only `r format(n_spp_both, big.mark = ',')` species included in both.  Here we examine differences in predicted species ranges between the two datasets, and find that these misalignments mainly result from divergent methodologies that introduce differing frequencies of commission and omission errors.  We illustrate the scientific and management implications of these differences by repeating two recent applications - an assessment of global biodiversity within the Ocean Health Index (Halpern et al. 2012, 2015) and a global analysis of gaps in coverage of marine protected areas (Klein et al. 2015) - and find significantly different estimates of the status of global biodiversity and effectiveness of conservation depending on how the two datasets were prioritized. Until a single, highly accurate dataset of global marine species ranges becomes available, understanding the implications of dataset differences for conservation planning and decision-making will be essential.
